Stage-Splitting a Modal Language

### 2014-06-05 paper/paper.tex
added a few words about the semantics

\documentclass{article}
\usepackage{amsmath}
\usepackage{bbm}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{amssymb}
\usepackage{mathabx}
\usepackage{float}
\usepackage{cite}
\usepackage{mathpartir}
\usepackage[left=3cm,top=3cm,right=4cm,nohead,bottom=3cm]{geometry}

\input {macros}

\title{\Large\textbf{General Stage Untangling with Applications to Algorithm Derivation}}
\author{Nicolas Feltman et al.}
\begin{document}
\maketitle

\input {grammars}
\input {semantics}
\input {splitting}

@@

Many computational tasks are premised on the notion of multiple domains, places and times at which computation occurs.  Yet programming languages traditionally view their execution environments as monolithic, specifying the What and the How, but not the When nor the Where.   As always, where the language falls short, a programmer instead must bear the burden.  Inevitably, this comes at the cost of expressiveness, safety, or both.  In this work, we present a language for specifying certain multi-domain computations, sacrificing neither the expressiveness nor the safety that programmers have come to expect in single-domain languages.

Consider the programming of a pipeline system.  Generally in such a system, data comes in to one processor, is transformed to an intermediate result, and then fed to later processors, possibly with some buffering, duplication, and reordering in between.  We can view an algorithm for such as system as one computation, split across two spatial domains (processors).  For a concrete example, we draw inspiration from graphics systems.  Lets say that we want to map a function $f : \mathrm{PixelCoord} \to \mathrm{Color}$ over every pixel on a 1024x768 screen.   Due to geometric coherence inherent in the problem, it's often advantageous to compute some information at the level of coarse 32x32 tiles (e.g. a list of the scene primitives which interact with that part of the screen) whose results can be used to accelerate the final per-pixel calculation.  In addition to yielding algorithmic benefits, this decomposition strategy maps well to parallel hardware.

Often when programming high-performance systems, a programmer will wish to specialize a bivariate function $f : \alpha \times \beta \to \gamma$ to one of its inputs, $x : \alpha$, while leaving the other, $y : \beta$, for later.  This need may arise because $x$ varies at a higher frequency than $y$, and we would wish to avoid repeated calculation. It may also be the case that $x$ is available at a time when computational resources are less expensive than when $y$ becomes available.  Here we say that the calculation is split accross two temporal domains.  Pulling another example from graphics, we consider a raytracer wherein $x : \mathtt{Geom}$ is a specification of the scene geometry (perhaps a list of triangles), $y : \mathtt{Ray}$ is a ray cast from some point in the scene, and $f$ computes the first intersection of the ray and the geometry.  In modern raytracers, we may cast millions of rays without varying the scene geometry, so it could be advantageous to precompute as much of the algorithm as possible based on only scene geometry.

In addition to the partitioning of a calculation across multiple domains, the examples above exhibit one further similarity: communication between domains is one-directional.  In the spatial case, information always flows from one processor to the next, but never backwards.  In the temporal case too, information made available at the second time cannot be observed earlier (indeed this seems to be a defining feature of time, generally).  For this reason, we refer to the domains as stages.  This property, that information flow is acyclic, informs the rest of this paper.

[Talk about how languages without inherent support for domains must instead tackle the problem by writing various passes/shaders. This is bad because it's not compositional.  For instance, I cannot write a function that talks about multiple stages at once.]

[Because of the acyclic nature of the problem, we can take algorithms specified in our multi-domain language and tease them apart, essentially recovering the old way of doing things.  This is called stage untangling.  It gets us all the safety and expression without messing up the rest of the compiler.]

[Talk about standard partial evaluation and pass separation, then place our work in the second, modulo binding time analysis.]

%\subsection {Program Generation} 
%A common solution is to write code that writes code.  Essentially, second-stage computations are represented by second-stage code, and second-stage code is a value that can be passed around and manipulated by first-stage code.  To use the terminology from above: this first stage code is a function that takes in $x$ and outputs second stage function, which itself takes in $y$ and outputs the final answer. Note that this is essentially a manual process, in that the programmer must know the function $f$ which is being specialized.  Using the equational notation of \cite{jones96}, where ``$\llbracket h \rrbracket~z$'' means the ``the code $h$ interpreted as a function and applied to input $z$," we say that a programmer defines a program generator $g$ that satisfies 
%\[
%\llbracket \llbracket g \rrbracket~x\rrbracket~y =\llbracket f \rrbracket~(x,y)
%\] 
%for all $x$ and $y$. When well supported by the language (\cite{devito13}, [some metaml citation]) this technique is often called {\em metaprogramming}.  Of course, even without first class support, one can represent the second stage code using strings or other data types.  Defined so broadly, program generation includes all compilers, where $f$ is a hypothetical interpretter, $g$ is the actual compiler, $x$ is a program in the source language, and $y$ is the input to $x$.  Program generation has also been used to implement the DFT example in the form of FFTW \cite{FFTW05}.  
%
%The benefit of the program generation technique is full control, even up to the ability to use domain-specific optimizations.  This comes at the cost of essentially being a manual operation.  
%
%\subsection {Partial Evaluation} 
%The idea of partial evaluation is to {\em automatically} specialize the code of $f$ to the supplied input, in a manner similar to standard evaluation.  In particular this means defining a program $pe$, called a partial evaluator, to which we can pass the implementation of a function $f$ along with $f$'s immediate input $x$, and get back a version of $f$ that is {\em specialized} to $x$.  This specialized version, also called a {\em residual}, is often denoted as $f_x$.  Equationally, we have
%\begin{align*}
%\llbracket pe\rrbracket~(f,x) &= f_x \\
%\llbracket f_x\rrbracket~y &= \llbracket f\rrbracket~(x,y)
%\end{align*}
% for all $f$, $x$, and $y$.  As seminally observed by \cite{futamura71}, partially evaluating an interpretter on a source program is equivalent to compiling that program to whatever language the interpretter was written in!  This has the benefit of being an automatic process, although the effective compiler is only as good as the partial evaluator.  Much of the subsequent work in partial evaluators has been with the purpose of chasing this goal.
%
%Partial evaluation is a more general technique than program generation, in that the partial evaluator is defined once and works for all $f$s that need to be specialized.  Of course, the partial evaluator has no domain knowledge of the program it is splitting, so it cannot be nearly as aggressive as a program generator might be.  The two techniques are similar in that they both require the first input $x$ to do their specializing action.
%
%\subsection {Pass Separation} 
%The final technique, pass separation, is conceptually the simplest.  The idea is to define a program called $ps$, for {\em pass separator}. It works by cleaving the function $f:\alpha \times \beta \to \gamma$ into two functions $f_1 : \alpha \to \tau$ and $f_2 : \beta \times \tau \to \gamma$ for some type $\tau$, where $f_1$ builds a data structure (of type $\tau$) from $x$, and $f_2$ consumes that data structure as well as $y$ to produce the standard ouput.  In analogy to partial evaluation, we call $f_2$ the {\em residual}.  Equationally, this is
%\begin{align*}
%\llbracket pe\rrbracket~f &= (f_1,f_2) \\
%\llbracket f_2\rrbracket~(\llbracket f_1\rrbracket~x,y) &= \llbracket f\rrbracket~(x,y)
%\end{align*}
%for all $f$, $x$, and $y$.  This technique, performed manually, is a common exercise for every programmer.  In can be considered the general form of common compiler optimizations such as loop hoisting, an example considered later.  Like partial evaluation, pass separation is an automatic operation, but unlike both partial evaluation and program generation, pass separation does not require the value of the first argument, $x$, to properly specialize $f$.  This makes pass separation the most widely applicable of the staging techniques, at the cost of being able to perform aggressive optimizations in the residual that might depend on the first argument. 
%
%In this work we present a pass separation algorithm for a fragment of ML.  Specifically, our formulation can separate terms containing sums and recursion, which have not appeared previously in the stage seperation literature.  We anticipate that the algorithm will also be able to also separate this language extended with with first-class functions, although this is left as future work.  Additionally, our presentation is more explicitly type-motivated than those that have come before.
%
%\section{Binding-Time Analysis and Stage Untangling}
%
%The pass separation problem can naturally be decomposed into two subproblems:
%\begin{itemize}
%\item Decide which parts of the definition of $f$ belong to which stage and produce an annotated version of $f$.
%\item Use the annotated version of $f$ to produce $f_1$ and $f_2$.  
%\end{itemize}
%
%An analogous form of this decomposition exists in the partial evaluation literature, wherein the first subproblem is known as {\em binding-time analysis}.  The second subproblem, in which the difference between pass separation and partial evaluation is more manifest, does not seem to yet have a name.  We call it {\em stage untangling}.  Unlike previous work in pass separation, we focuses entirely on the stage untangling problem by assuming that $f$ is given to us in an annotated form. 
%
%Previous research has observed that the pass separation problem is inherently ambiguous in that the equations defined above do not fully specify the definition of $pe$.  That is, there can exist multiple ways to partition $f$ into $f_1$ and $f_2$ that satisfy $\llbracket f_2\rrbracket~(\llbracket f_1\rrbracket~x,y) = \llbracket f\rrbracket~(x,y)$.  For a trivial example, we can set $f_1$ to be the identity and $f_2$ to be $f$. Fortunately, this ambiguity is contained entirely within the binding-time analysis portion of pass separation; stage untangling is entirely determined.

\section{A Simple Two-Stage Language}

We'll start by analyzing a simple two-stage language called \ellStaged.  We say it is simple in that it supports boolean, integer, and product types, as well as let bindings, if expressions, and various primitive operations.  Notably, we leave out functions, fixed point operators, and full sum types, which will be added later.

The grammar for types, terms, values, and contexts in \ellStaged are given in Figure \ref{fig:coreSyntax}.  The first thing to notice is that stage $\bbone$ and stage $\bbtwo$ code are syntactically seperated at the type, term, and value levels, despite heavy overlap.  It would be tempting to merge the two stages into one syntactic class and maintain the stage invariants with the typing judgements.  Such a trick would reduce the total number of rules presented later, but at the cost of making those rules type derivation-directed, rather than syntax directed.  Alternatively, we chose to have one context merged accross both stages, rather than one context for each stage.  The intended effect of this decision is for bindings in one stage to shadow binding to the same variables in the other stage.  The reasons for this will become clear later.

At the type and value levels, stage $\bbone$ can refer to stage $\bbtwo$, but stage $\bbtwo$ cannot refer back to stage $\bbone$.  Alternatively, the two stages are mutually recursive at the term level.  These two patterns together fulfill the promise of {\em mixed code, staged evalutation}.

In the rest of this document, I will tend to use $\{A,B,C\}$ for types, $\{v,u,w,t\}$ for values, $\{x,y\}$ for variables, $e$ for terms, $i$ for integers, and $b$ for booleans.

\subsection{Static Semantics}

The static semantics for \ellStaged~is given in Figure \ref{fig:coreStatics}.  Modulo the two-stage restriction and current absence of functions, it's based off of the system in \cite{davies96}, which was inspired by temporal logic (the source of our $\fut$ modality).  The static semantics comprises two judgements.  Those are:

\begin{center}
\begin{tabular}{|l|l|} \hline
Name & Pattern \\ \hline
Typing at $\bbone$ & $\typesone[\contextot]{\expro}{\typeo}$ \\  \hline
Typing at $\bbtwo$ & $\typestwo[\contextot]{\exprt}{\typet}$ \\ \hline
\end{tabular}
\end{center}

We see that the typing judgements for both stages are each what one would expect for an unstaged language.  That is, there are introduction and elimination forms for base types and products at both stage $\bbone$ and stage $\bbtwo$.  Furthermore, these rules preserve stage.  For these reasons, we can think of the base type and product features as essentially being orthogonal to the staging features.  The terms $\next$ and $\prev$, which are the only terms that allow us to change stage, are respectively the introduction and elimination forms for $\fut$.

\subsection{Erasure Dynamic Semantics}

The dynamics semantics for \ellStaged~is given in Figure \ref{fig:erasureSemantics} in big-step style.  Predictably, it comprises two judgements:

\begin{center}
\begin{tabular}{|l|l|} \hline
Name & Pattern \\ \hline
Reduction at $\bbone$ & $\reduceone{\expro}{\valo}$ \\  \hline
Reduction at $\bbtwo$ & $\reducetwo{\exprt}{\valt}$ \\ \hline
\end{tabular}
\end{center}

The primary goal for this semantics was simplicity.  This was achieved in that, like the static semantics, the judgements for each level are extremely similar to those for an unstaged language.  This simplicity came at the cost of some descriptiveness, however.  The inadequacy of this semantics will become apparent shortly.

First, we discuss a few desirable theorems.  Chiefly among them is type preservation (I haven't defined the judgements for the types of values, but it should be rather clear).  As will be the pattern for many theorems to come, our inductive hypothesis comprises two mutually dependent lemmas, one for stage $\bbone$, and the other for stage $\bbtwo$:
\begin{center}
\begin{tabular}{l}
If $\typesone [\emptyC]{e}{A}$ \\
and $\reduceone{e}{v}$ \\
then $v : A$
\end{tabular}
~~~
\begin{tabular}{l}
If $\typestwo [\emptyC]{e}{A}$ \\
and $\reducetwo{e}{v}$ \\
then $v : A$
\end{tabular}
\end{center}
[Proof not done, but it looks trivial in every case. [Wait no, substitutions kill me again.]]

We also want some formalization of the idea that stage two computation cannot affect stage one computation.  At a minimum, the following lemma should be true [this needs lots of work],

\begin{center}
\begin{tabular}{l}
If $\exists v\in \valt$ s.t. $\reduceone {[v/x]e}{u}$ \\
then $\forall v\in \valt$ s.t. $\reduceone {[v/x]e}{u'}$, $u \simeq u'$
\end{tabular}
\end{center}

As it turns out, this is not true! [Example forthcoming.]

\section{The Stage-Splitting Transformation}

We now cover splitting, the process by which mixed code is teased apart into two individual chunks of code, one for each stage.  To begin, we first introduce the target lagnuage (for both stage), \ellTarget.   The grammar for \ellTarget~is given in figure \ref{fig:baseSyntax}.  We see that \ellTarget~is (as one would hope) a traditional single-stage language, with all the same basetypes as \ellStaged, in addition to products, if expressions, and full sums (injections and case expressions).  The presence of this last feature is a bit curious, since \ellStaged~did not support full sums.  The necessity for sums should become apparent in time.  The static and dynamic semantics of \ellTarget~are not given, but should be obvious.  They are given by the judgements:

\begin{center}
\begin{tabular}{|l|l|} \hline
Name & Pattern \\ \hline
Typing & $\types{\expr}{\val}$ \\  \hline
Reduction & $\reduce{\expr}{\val}$ \\ \hline
\end{tabular}
\end{center}

The splitting judgements come in seven parts, two each for types, terms, and values, and one for contexts.  In tabular form, they are:
\begin{center}
\begin{tabular}{|l|l|} \hline
Name & Pattern \\ \hline
Value Splitting at $\bbone$& $\valo \vsplito [\val,\val]$ \\  \hline
Value Splitting at $\bbtwo$& $\valt \vsplits \val$ \\  \hline
Type Splitting at $\bbone$& $\typeo \tsplito [\type,\type]$ \\  \hline
Type Splitting at $\bbtwo$& $\typet \tsplits \type$ \\  \hline
Context Splitting & $\contextot \csplit [\context,\context]$ \\  \hline
Term Splitting at $\bbone$& $\splitone{\expro}{\typeo}{\expr,\var.\expr}$ \\  \hline
Term Splitting at $\bbtwo$& $\splittwo{\exprt}{\typet}{\expr,\var.\expr}$ \\  \hline
\end{tabular}
\end{center}

[...more...]

\subsection {Splitting Theorems}

\subsubsection{It preserves types}

The theorem we want to show has two parts, one for the $\bbone$ code, and one for the $\bbtwo$ code.  
\begin{center}
\begin{tabular}{l}
If $\splitone {e}{A}{c,l.r}$ \\
then $\Gamma \csplit [\Gamma_1;\Gamma_2]$ \\
and $A \tsplito [A_1,A_2]$ \\
and $ \types [\Gamma_1]{c} {A_1 \times \tau}$ \\
and $ \types [\Gamma_2, p:\tau]{r} {A_2}$ 
\end{tabular}
~~~
\begin{tabular}{l}
If $\splittwo {e}{A}{p,l.r}$ \\
then $\Gamma \csplit [\Gamma_1;\Gamma_2]$ \\
and $A \tsplits A'$ \\
and $ \types [\Gamma_1]{p} {\tau}$ \\
and $ \types [\Gamma_2, p:\tau]{r} {A'}$ 
\end{tabular}
\end{center}

\subsubsection{It preserves meaning}

The theorem we want to show has two parts, one for the $\bbone$ code, and one for the $\bbtwo$ code.  
\begin{center}
\begin{tabular}{l}
If $\splitone [\emptyC]{e}{A}{c,l.r}$ \\
and $\rsone{e}{v}{q}$ \\
and $\reducetwo{q}{w}$ \\
then $\reduce{c}{\valprod{v}{u}}$ \\
and $\reduce{[u/l]r}{w}$
\end{tabular}
~~~
\begin{tabular}{l}
If $\splittwo[\emptyC]{e}{B}{p,l.r}$ \\
and $\rstwo{e}{q}$ \\
and $\reducetwo{q}{w}$ \\
then $\reduce{p}{u}$ \\
and $\reduce{[u/l]r}{w}$ 
\end{tabular}
\end{center}

\section {Adding Full Sums and Functions}

In this section, we consider adding two more features to our source language: sums and functions.  We will call this augmented language \lamStaged.  The grammar for \lamStaged is given in \ref{fig:lamStagedSyntax}.

\section{Related Work in Stage Separation}

Our starting language, $\lambda^{12}$, is essentially a two-stage version of $\lambda^\bigcirc$ from \cite{davies96}, but beefed up with products, sums, and fixed-points.  That work motivates $\lambda^\bigcirc$ as an image of linear temporal logic under the Curry-Howard correspondence, and it makes the argument that $\lambda^\bigcirc$ is a good model for binding time analysis.  Although $\lambda^\bigcirc$ was designed with partial evaluation in mind, it turns out to be a good model for pass separation as well.  In contrast, \cite{davies01} presents $\lambda^\Box$, which is based off of branching temporal logic.  As that work shows, $\lambda^\Box$ is a good system for program generation, although we find the closed-code requirement to be more restricting than we need.

Pass seperation was introduced in \cite{jorring}.  They focused on motivating pass separation as a technique for compiler generation, and hinted at wider applicability.  Their approach was not mechanized or automated in any sense, and their examples were separated by hand.  We believe that out work represents the fulfillment of their prediction that ``the [pass separation] approach will elude full automation for some time."

\cite{hannan94} also uses a pass separation technique for generating compiler/evaluator pairs from interpretters specified as term-rewrite systems.  This heavy restriction on the form of the input prevents generalization of their style of pass separation.  [I don't know enough about term-rewrite systems yet to be able to comment further.]  Their method, while amenable to mechanization, was not actually implemented.

\cite{knoblock} represents the first attempt at fully automatic pass separation for code approaching general-purpose.  In particular, they implemented pass separation for a C-like shading language including basic arithmetic and if statements.  The main goal of their work, like ours, was to minimize recomputation. That said, their starting language was sufficiently restricted for recursive boundary types to be inexpressible.  Given that memory is at a premium in shading languages, so much of that work was also focused on minimizing the memory-footprint of the boundary type.

\section{Appendix}

\subsection {Equivalence of Diagonal Evaluation and Splitting (for \ellStaged)}

We proceed by cases of the term splitting judgement.

\begin{itemize}
% INT, BOOL, UNIT
\item {\bf Int.} Suppose that
\begin{enumerate}
\item $\splitone{i}{\rm int}{(i,()),\_.()}$
\item $\rsone{i}{v}{q}$
\item $\reducetwo{q}{w}$
\end{enumerate}
By inversion of reduction relations, we know
\begin{enumerate}
\item $v$ has the form $i$
\item $q$ has the form $()$
\item $w$ has the form $()$
\end{enumerate}
It is directly provable that,
\begin{enumerate}
\item $\reduce{(i,())}{\valprod{i}{()}}$
\item $\reduce{[()/\_]()}{()}$
\end{enumerate}
Which is what we need to show.

\item {\bf Boolean.} Same as int.
\item {\bf Unit.} Same as int.

>>>>

\documentclass{sigplanconf}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}
%\usepackage{amssymb}
\usepackage{bbm}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{mathpartir}
\usepackage{mathabx}
\usepackage{listings}

\begin{document}

\input {macros}
\input {grammars}
\input {semantics}
\input {splitting}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
\copyrightyear{20yy} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
\doi{nnnnnnn.nnnnnnn}

% Uncomment one of the following two, if you are not going for the 
% traditional copyright transfer agreement.

%\exclusivelicense                % ACM gets exclusive license to publish, 
                                  % you retain copyright

%\permissiontopublish             % ACM gets nonexclusive license to publish
                                  % (paid open-access papers, 
                                  % short abstracts)

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Stage-Splitting a Modal Language}

\authorinfo{Name1}
           {Affiliation1}
           {Email1}
\authorinfo{Name2\and Name3}
           {Affiliation2/3}
           {Email2/3}

\maketitle

\begin{abstract}
This is the text of the abstract.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

% general terms are not compulsory anymore, 
% you may leave them out
\terms
term1, term2

\keywords
keyword1, keyword2

@@

Staged computation has existed as a programming technique for over three decades.  Jorring et al. (\cite{jorring86}) identify three classes of staging techniques: meta-programming, partial evaluation, and stage-splitting.  The first two of these have received significantly more attention than the third.

We call these techniques {\em staged} in that part of their input comes in the first stage, and part comes in the second stage.  Likewise, some computation occurs at each stage as well.

Countless meta-programming systems exist (...twelve thousand citations...\cite{devito13}), and their background theory and type-systems are well understood (\cite{davies01}).  Partial evaluation, too, is well-understood.  Partial evaluation systems exist... . This paper explores both theory and applications for stage-splitting.

\section{Stage-Splitting Definition and Comparison to Partial Evaluation}

First, we review the definition of partial evaluation.  Informally, a partial evaluator takes the code for a function $f$, as well as the first-stage input $x$ to that function, and produces the code for a version of that function {\em specialized} to the first input, often called $f_x$.  This $f_x$ function can then be evaluated with the second-stage input to produce the same final answer that $f$ would have.  The goal of the process is that $f_x$ should be cheaper to evaluate than $f$, although this can't be guaranteed for all inputs.  We now state this theorem more formally: a partial evaluator is some function $p$ such that,
\[
	\forall f,x. \exists f_x. [p(f,x) = f_x \text{ and } \forall y.\llbracket f \rrbracket(x,y)=\llbracket f_x \rrbracket(y)]
\]
where (borrowing notation from \cite{jones96}), $\llbracket f \rrbracket$ means the mathematical function corresponding to the code given by $f$.

Informally, we define stage-splitting to be the process of taking some function $f$ into two other functions, $f_1$ and $f_2$, where $f_1$ computes a partial result from the first-stage input, and $f_2$ uses that partial result and the second-stage input to compute a final result which is the same as if we had just run the original $f$ on both inputs.  Again, more formally, a stage-splitter is some $s$ such that,
\[
	\forall f. \exists f_1,f_2. [s(f) = (f_1,f_2) \text{ and } 
	\forall x,y.\llbracket f \rrbracket(x,y)=\llbracket f_2 \rrbracket(\llbracket f_1 \rrbracket(x),y)]
\]

We first discuss a few similarities between partial-evaluation and stage-splitting.  First off, both techniques have the same form of input, namely a bivariate function where the first input comes at stage one, and the second input comes at stage two.

Again in both cases, the governing equations are too weak to fully determine the definitions of $p$ and $s$.  Indeed, both admit completely trivial definitions.  Consider the stage-splitter which always returns the identity for $f_1$ and $f$ for $f_2$, or analogously the partial evaluator which always returns an $f_x$ that just closes over the input $x$ and internally calls $f$ once $y$ is available. The ambiguity of these equations (modulo standard program equivalence of the outputs) can be resolved by adding annotations to $f$ to clearly specify the parts of the computation that are first stage and the parts that are second stage.  Later, we show that the same annotations suffice for both partial evaluation and stage-splitting.

The differences between stage-splitting and partial evaluation are likewise evident from these governing equations.  For instance in partial evaluation, the existential $f_x$ depends on $x$, which means that the partial evaluator cannot be run until $x$ is known.  Moreover, if one wishes to specialize $f$ for multiple $x$'s, then the partial evaluator must be run several times.  Depending on the use case and cost of partial evaluation, this may be prohibitively expensive.  Alternatively, a stage-splitter need only be run once, and this can be done entirely before any $x$ is known.

\subsection{Partial Evaluator from Stage-Splitter}

We can recover a valid partial evaluator from a stage-splitter by stage-splitting the input function $f$ into $f_1$ and $f_2$, computing $\llbracket f_1 \rrbracket(x)$ to obtain $\bar x$, and then returning an $f_x$ such that $\llbracket f_x \rrbracket(y) = \llbracket f_2 \rrbracket(\bar x, y)$.  Note that this does not mean that stage-splitting is a strict generalization of partial evaluation.  In practice, partial evaluators easily perform optimizations (such as branch elimination, discussed later) which are beyond the scope of stage-splitting, and would require further technology than has been developed here.  It is best to think of stage-splitting as simply the first half of partial evaluation, where the back half is an optimizer. [Might be able to come up with a futamura projection-like statement here, which would be really really really cool.]

\subsection{Stage-Splitter from Partial Evaluator}

Likewise, we can easily recover a stage-splitter from a partial evaluator.  If $p$ is a valid partial evaluator, then we can define a stage splitter $s$ such that $s(f)=(f_1,f_2)$, where
\begin{align*}
[f_1](x) &= p (f,x) \\
[f_2](l,y) &= [l] (y)
\end{align*}
This implicitly requires that the languages in which $f_1$ and $f_2$ are expressed are strong enough to write a partial evaluator, but that is the case in this paper.  A stage-splitter defined this way leaves much to be desired.  Firstly, partial evaluation of $f$ may be too expensive for the context in which $f_1$ needs to run.  Additionally, the intermediate data structure created this way may be much larger than necessary, as it would contain all of the residual code.

\section{A Two-Stage Modal Language}

Introduce the next and prev concepts, along with typesystem.  Introduce binding time analysis here, and explain that we don't care about it.  Show some examples.  Introduce a hold operation.

\section{Diagonal Semantics}

We desire a semantics that meets the following goals,

\begin{enumerate}
\item In absence of $\next$ expressions, it should be equivalent to standard call-by-value evaluation.
\item $\next$ and $\prev$ should be inverses of the other.
\item All of the first stage work should be completed before second stage work.
\item The result for a stage-2 expression should be the same as if all of the $\next$s and $\prev$s were interpreted as the identity function.
\end{enumerate}

In this section, we introduce a dynamic semantics for our language.  Although our type system is virtually identical to that in \cite{davies96}, the semantics is different in terms of cost and termination behavior.  The semantics comprises the following judgments:

\begin{center}
\begin{tabular}{|c|p{2.5cm}|p{2.5cm}|} \hline
Judgment & Reads & Conditions \\ \hline 
$\isvalone e$ & ``$e$ is a $\bbone$-value under context $\Xi$'' & ... \\ \hline 
$\Xi \vdash \diaonesub$ & ``$e$ evaluates to future table $\gamma$ and value $v$'' 
& $\typesone e A$ 
	\newline $\isvalone v$ 
	\newline ... \\ \hline 
$\diatwosub$ & ``$e$ speculates to $q$'' & ... \\ \hline 
\end{tabular}
\end{center}

\subsection{Adherence to Goals}

We meet our first and second goals (show examples).

That we meet our third goal is clear from just the structure of our judgments (...).

Under the strictest interpretation, we do not meet our third goal.  Consider the following examples (show speculating down an if).

But as it turns out, a semantics which had the erasure behavior would not meet goal 3! (explain using the same example)

\section{Splitting Algorithm}

[Present the splitting judgement.  Give statements of type and value correctness for splitting.  Give all of the splitting rules.  Talk through a few of them.]

The theorem that we want to be true is...

%\begin{center}
%\begin{tabular}{l}
%If $\splittwo{e}{A}{p,l.r}$ \\
%and $\reducetwo{e}{v}$ \\
%then $\reduce{p}{u}$ \\
%and $\reduce{[u/l]r}{w}$ \\
%and $v = w$ 
%\end{tabular}
%\end{center}

%\begin{center}
%\begin{tabular}{l}
%If $\splitone {e}{A}{c,l.r}$ \\
%and $\colone {\gamma_1}{\Gamma}$ \\
%and $\coltwo {\gamma_2}{\Gamma}$\\
%and $\rsone{\gamma_1(e)}{v}{q}$ \\
%and $\reducetwo{\gamma_2(q)}{w}$ \\
%then $\reduce{|\gamma_1|(c)}{\valprod{v}{u}}$ \\
%and $\reduce{[u/l]|\gamma_2|(r)}{w}$
%\end{tabular}
%~~~
%\begin{tabular}{l}
%If $\splittwo{e}{B}{p,l.r}$ \\
%and $\colone {\gamma_1}{\Gamma}$ \\
%and $\coltwo {\gamma_2}{\Gamma}$\\
%and $\rstwo{\gamma_1(e)}{q}$ \\
%and $\reducetwo{\gamma_2(q)}{w}$ \\
%then $\reduce{|\gamma_1|(p)}{u}$ \\
%and $\reduce{[u/l]|\gamma_2|(r)}{w}$ 
%\end{tabular}
%\end{center}

\section{Examples for Staged Pipelines}

Give the gist of one-to-one pipeline example (like client/server).
Then talk about a one-to-many pipeline.
Then talk about a many-to-one pipeline like spark.  It clear how to target something like this for known base types on the boundary, and for product types.  But sums on the boundary are hard!  We leave many-to-one as future-work.

\section{Examples of Algorithm Derivation}

Fast exponent example.

\lstset{language=ML, columns=fullflexible, basicstyle=\ttfamily, tabsize=4, escapeinside={"*}{*"}}
\begin{lstlisting} 
let exp (b : $int, e : int) : $int =
	if e == 0 then
		next{1}
	else if (e mod 2) == 0 then
		next{let x = prev{exp(b,e/2)} in x*x}
	else
		next{prev{b} * prev{exp (b,e-1)}}		
\end{lstlisting}

splits into

\begin{lstlisting} 
let exp (b, e) =
	((), roll (
		if e == 0 then
			inL ()
		else 
			inR (
				if (e mod 2) == 0 then
					inL (#2 (exp (b,e/2)))
				else
					inR (#2 (exp (b,e-1)))
			)
	))
\end{lstlisting}

and

\begin{lstlisting} 
let exp ((b, e), p) =
	case unroll p of
	  () => 1
	| d  =>
		case d of
		  r => let x = exp ((b,()),r) in x*x
		| r => b * exp ((b,()),r)
\end{lstlisting}

Quickselect example.

\begin{lstlisting} 
let qs (l : "*$\mathtt{\mu \alpha. }$*"() + "*$\mathtt{int*\alpha}$*", i: $int) = 
	case unroll l of
	  () => next {0}
	| (h,t) => 
		let (left,right,n) = partition h t in
		next{
			let n = prev{hold n} in
			case compare prev{i} n of
			  () (*LT*) => prev {qs left i}
			| () (*EQ*) => hold h
			| () (*GT*) => 
				prev {qs right next{prev{i}-n-1}}
		}	
\end{lstlisting}

Things to try: an interpreter which, partially evaluated, does cps or something.

For each of these examples, talk about what partial evaluation would do and why that might be bad.

\section{Related Work}

Our stage-splitting algorithm was first suggested in \cite{jorring86} under the name {\em pass separation}.  They essentially proposed that a function $f$ could be split into two others, $f_1$ and $f_2$, such that $f(x,y)=f_2(f_1(x),y)$.  They did not distinguish binding time analysis from stage splitting, and so pass separation inherits the former's ambiguity.  The main goal of \cite{jorring86} was to motivate pass separation and other staging transformations as a powerful way to think about compilation and optimization.  Accordingly, their approach was entirely informal, with no implementation realized.  Moreover, they predicted that ``the [pass separation] approach will elude full automation for some time."

Implementations of the stage-splitting algorithm have appeared in the literature exclusively (and coincidentally) in the context of graphics pipelines.  The first of these (\cite{knoblock96}) uses a binding time analysis to separate those parts of graphics shaders that are input-invariant from those which are not, and then uses a stage splitting algorithm to factor that into two shaders, thereby minimizing recomputation.  Their shaders are written in a C-like language with basic arithmetic and if statements.  Although their analysis does not give an explicit account of the type-level behavior of the splitting algorithm, it effectively can synthesize product and sum boundary type.

Like the previous example, the Spark language (\cite{sparkThesis}) uses staging to minimize recomputation in real-time rendering applications.  But instead of using a binding-time analysis, Spark allows the programmer to manually target stages of the graphics pipeline.  Since the modern graphics pipeline is inherently a many-to-one system, this is difficult to reconcile with sum types on the boundary.  Fortunately, Spark has a set of syntactic restrictions that prevent sum boundary types.  Spark does not clearly identify this conflict, but the authors did note that first-stage if statements were difficult to provide meaning to [need quote].

[RTSL and SH]

[Discuss Yong's recent paper here.  It does some pretty sophisticated binding time analysis, with a somewhat straightforward splitting after that.  They have the same many-to-one use case as Spark, but syntactic restrictions prevent sum types on the boundary, sort of.  If we wanted to faithfully represent their system in ours, we would need some mechanisms for abstraction over stage, which we do not have.]

Davies (\cite{davies96}) explored the connection between linear temporal logic and its corresponding type system [circle](which we adapted into [circle sub 2]), and showed the equivalence between [circle] and existing systems for binding time analysis. That work provided $\beta$ and $\eta$ rules for the next and prev operators, but did not consider a full dynamic semantics for the whole language. Whereas [name of our type system] is appropriate for stage-splitting and partial evaluation, \cite{davies01} provides a similar system, [square], that is appropriate for meta-programming.  The main difference is that terms inside a [prev] operator do not see any stage-2 bindings declared outside of it.  They note that where [circle] corresponds to the non-branching temporal logic, [square] corresponds to the branching version.

[Meta-ML eases off on this restriction but does not (I think?) eliminate it.]

[What's going on with names and necessity?]

[Our work bears a lot of similarity to ML5, which also uses a modal type system.  The difference is that we target stages systems (each stage talks to the next), whereas they target distributed ones (all stages talk to all others). The type systems reflect this directly in the world accessibility relation.  There might be some analogue of stage-splitting in the ML5 work, but I have not yet isolated it (might be buried in CPS conversion).]

\appendix
\section{Appendix Title}

This is the text of the appendix, if you need one.

\acks

Acknowledgments, if needed.

% We recommend abbrvnat bibliography style.

\bibliography{paper}
\bibliographystyle{abbrvnat}
